Manuscript_draft

### 2021-03-16 manuscript_draft.Rmd
moved growth dominance statistic seciton around

---
title: "Manuscript_draft"
author: "Michael Jull and Hardy Griesbauer"
date: "10/12/2020"
output:
  word_document:
    reference_docx: wordTemplate2.docx
  html_document:
    df_print: paged
csl: cjfr.csl
bibliography: ep1162.bib

---

@@

- Great: Very clear and straightforward. i.e. - log-transformed treatment (initial BA) was a very statistically significant predictor of tree response across dbh classes for both Sx and Bl.
- Don't know what all the random effects outputs are. I am sure they are meaningful, but I just dont know what they mean.
- Number of observations of 510: Should ensure it is clear that only trees > 7.5 cm dbh are counted.
- Would be good to define the dbh class limits somewhere. In Appendix 6 table, we describe the class mid-point.
- For Appendix 6, I assume that the null hypothesis for the ratio is = 1.

@@

- I am thinking that mortality should / could be analysed as a frequency response - i.e. - expected vs actual distribution of mortality by size class and/or species. e.g. a Chi-squared or allied type of analysis. Null hypothesis is that mortality of stems is randomly distributed across different sizes and species of trees. H1 is that it is non-random.
- I'd suggest that we don't compare total mortality (e.g # of stems) across treatment units, but rather frequency relative to initial number of stems (i.e. - at beginning of observation period). That way we normalize mortality rates relative to initial stem densities. Otherwise, results could be mis-leading - e.g. Control treatments could have more mortality in overall sph, simply because they have more initial stems.
- ACTION ITEM FOR ME:  I need to check my original datasets and data sorting iterations to see how I handled mortality observations for the 1992 and 1994 plot remeasurements.
- Note for you:  All of the mortality you've analysed and identified so far is mortality that is incremental following 92/94, so it is still solid analysis data, just with that qualification. I will clarify the 92/94 mortalilty observations for the Methods section; we did observe dead trees at these times, but these mortality pre-dated our treatments and/or first measurements. Anyway, probably not a big deal, but this needs examination and clarification to explain this. If necessary, I will quantify observations of dead trees observed on these years, because they are actually > 0.

Testing for density-dependence of mortality responses... some ideas or hypotheses..  Note that f (___) means "a function of"...

  *   MORT RATE of a size class = f (Sum of BA in that plot > than that size class).
Would assume asymmetric competition where only bigger trees exert competition on smaller trees.
  *   MORT RATE of a size class = f ( Sum of BA in that plot that is either > or < that size class).
Assumes symmetric competition where both big and small trees exert competition on their neighbors.

@@

- See notes above about use of term "units" vs treatment (or similar).
- Clarify effects of treatment and/or species on growth rate vs standing vol / BA at each measurement period.
- A Figure 7b or equivalent could visually present changes in growth rate over time. (e.g. previous graphs we've developed)
- Under Observations, # 2 - the 2nd sentence needs work to edit and clarify.

@@

Last winter, as well as keeping reference copies of the initial “raw” dataset following data entry and related stuff, I kept copies each stage of data cleaning and prep (about 10 iterations).

We did have “dead tree” observations in the first ~ 2 year of monitoring. The key thing to see is : if I can determine is when the trees were dead, and/or when death was initiated, and why.  
Timing of tree death and observations of causal factors of tree death in these first two years will help inform our interpretation and data handling of these observations.

If trees were dead or moribund at the first “initial” observation or two (as soon after the logging as possible), this is – arguably -  different than mortality that occurs over time in response to the basal area treatment and/or could be considered to be density- or treatment-dependent in some way.

The 1992 data was in mid-March really early after logging, and before snowmelt (in retrospect not a great idea but we had resources available before the end of the fiscal year at the time). The 1994 data was from when the plots were re-visited under snow-free conditions. We did a thinning and sanitation culling of trees in the treatment units and PSP’s in 1994 to remove logging-damaged trees and most deciduous, to bring the units and PSP’s down to their final treatment BA. 1994 remeasurements were completed after that.

@@

1.	The very-early observed mortality may be a result of pre-existing stand conditions not the created, treatment-induced stand conditions. “stochastic” vs potentially treatment induced. 
2.	It can be argued that trees dead at time of treatment or very soon  thereafter were never taking up growing space in the post-harvest stand, and therefore would not contribute to the initial “green” basal area going forward. 
3.	We do have 1992 data in most plots, but it was 1994 when the BA in each trmt unit and embedded PSP was finalized. 1994 is the most homogeneous start point for all PSP’s.

@@

Trees that were dead at the time of the 1992 PSP establishment were not tagged or recorded. The rationale for this was that these trees were dead and therefore not taking up up growing space in the stands. Dead trees in 1992 would have been either snags / dead trees pre-existing the harvest, or trees fatally damaged or killed by logging activities (e.g. – trees with stems snapped and crowns broken off, no live limbs) or trees smashed to the ground by felled trees).

@@

EP 1162 treatment units, including the PSP area contained therein, were assessed in Spring 1994 to review stand density / basal area, and stand condition, in order to:  
a)	identify and mark-to-cut residual trees that were considered too badly scarred or damaged to constitute good growing stock for trial purposes, 
b)	where needed, to lower the initial BA closer to the target level, 
c)	to either fell or girdle birch trees, especially in the PSP’s themselves, but usually also within the TU as a whole, and 
d)	to fell any dead, dangerous, and/or moribund trees within the treatment units and PSP’s. 
Generally most if not all of these trees felled in the sanitation felling were understory or subdominant trees, and occasionally a lower codominant tree. Tagged trees felled by the sanitation felling were documented on field sheets and our Excel datasets, so we can retrieve the species and size of these trees as needed. However, sanitation-felled trees included both live, moribund, and dead trees – generally all poorer quality or vigor though.

After felling, the felled trees were bucked to a 2 to 3 m length so as to lay these felled trees to as close to the ground as possibly, maximize decay, and minimize physical obstructions within the TU’s and PSP’s.

@@

We don’t, therefore, have data on mortality between 1992 and 1994, though my recollection is that it was minor in nature.

@@

“The difference in basal area between 1992 and 1994 is a result of basal area reductions through the sanitation felling of some stems and some minor natural mortality, combined with basal area increases through the growth of remaining live stems.”

A review of the various plots indicates that between 1992 and 1994, some PSP’s decreased in BA (as sanitation removals and minor mortality exceeded growth of live stems), while others increased in BA as tree growth exceeded BA removals.

  *   Is there any change in which sizes or size classes of trees die as stand volume increases? - e.g. is competition asymmetrical or symmetrical? Currently the y axis is total number of trees, I believe.
  *   Not sure if we have enough sample size of trees to look at species dynamics (Sx vs Bl).
  *   I am wondering if the same relationship that appears between volume and tree mortality, is mirrored in the relationship between BA and tree mortality. A bit more on this below...

@@

To test the hypothesis that tree-level radial growth response over the 1994-2019 period varied with interactions between (i) tree diameter at the time of final treatment (1994); (ii) treatment; and (iii) species, we fit linear mixed effects models with log-transformed 1994-2019 diameter increment as the response variable, and treatment, species and log-transformed 1992 diameter as interacting fixed effects. To account for plot-level productivity differences, we included plot as a random intercept effect.  Models were fit using the 'lme4' package for R [@].  We also fit a similar model with the 1994-2019 diameter increment expressed as a percentage of 1994 diameter.  Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.  Variance explained for each model are reported using the approach described in Nakagawa et al. [-@Nakagawa2013], as implemented in the MuMIn package for R [@barton2009mumin].  Pairwise contrasts were performed using the 'emmeans' package for R [@] and Tukey HSD method of *p* value adjustment was used to control for multiplicity.  We report all significant results at p<0.05.

@@

#### Stand-level volume and basal area estimates
Total and merchantable volumes were estimated for trees using species- and region-specific equations [@nigh2016] based on tree height and DBH.  Consistent with forest management practices in British Columbia, we estimated volumes for trees meeting a minimum diameter utilization limit of 18.5cm DBH.  For trees lacking a tree height measurement, we estimated height by fitting a single diameter-height model developed with a linear mixed effects model, with log-transformed DBH as a fixed effect, and tree nested within plot as random intercepts.  Trees with broken tops were omitted from diameter-height models. We summed tree-level estimates of volume and basal area to generate stand-level estimates for each plot. To test the hypothesis that changes in stand attributes over 27 years varied by treatment, we fit separate linear models with stand total volume, basal area, and density as response variables, and the interaction between treatment and year as the independent variable.

@@

The tree-level basal area increment model indicated that interactions between log-transformed tree size, species and treatment explained `r round(r.squaredGLMM(fitBA)[2]*100,1)`% of the variation in log-transformed 1994-2019 tree-level BAI, and `r round(r.squaredGLMM(fitBApct)[2]*100,1)`% of the variation in basal area increment as a percentage of tree size in 1994 (p<0.0001 for both, `r tableNums("fitBA.table",display="cite")`).  Estimated marginal means from the model showed that between species and among treatments, basal area increment increased with tree size, meaning that larger trees increased basal area more than smaller trees (`r figNums("treeGrowth.plot",display="cite")`), although smaller trees increased their size proportionally more than large trees (`r figNums("treeGrowthPct.plot",display="cite")`).  Species contrasts showed that spruce trees increased  basal area significantly more than than fir in the high-removal and control treatment units, whereas basal growth between the two species was close to equal in the low-removal unit (`r appNums("sppContrast.table",display="cite")`).  Within-species contrasts showed that spruce basal area increment in the high-removal unit exceeded spruce growth in both other units (`r appNums("trtContrast.table",display="cite")`), whereas fir basal area increment in the high- and low-removal units was higher than the control unit, and there were no significant differences between the two harvested units (`r appNums("trtContrast.table",display="cite")`).

@@

### Tree height modeling
Tree heights had a positive nonlinear relationship with diameter, for both species and across all time periods (`r appNums("heightDiam.plot",display="cite")`).  We originally developed separate models for each species and time period, however, a comparison of height-diameter models showed that intercepts and slopes did not vary significantly (i.e., p>0.05) between species or time periods (not shown).  Therefore, we developed a single height-diameter linear mixed-effects model using `r X %>% filter(Species %in% c("Fir","Spruce")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset. This model explained `r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in log-transformed heights, with an intercept of `r round(fixef(htDiamFit)[1],3)` and slope coefficient of  `r round(fixef(htDiamFit)[2],3)` * log-transformed DBH (not shown).  Residual and predicted vs actual height plots were assessed to ensure goodness of fit (`r appNums("htModelFit.plot",display="cite")`). Using this model, we generated height estimates for trees without height measurements, and then estimated tree total volume as per Nigh [-@nigh2016].

@@

### Rate of change from 1994 to 2019
All treatment units, including the control, increased volume, basal area and stand density over the 1994-2019 period (`r figNums("standPlots",display="cite")` and `r appNums("standPlot.table",display="cite")`).  Mean annual basal area and volume increment peaked during the 1997-2009 period in the two harvested treatments, and peaked in the 1994-1997 period in the control treatment unit (`r figNums("perInc.plot",display="cite")`).  From 1994 to 2019, estimated marginal means of basal area increment in the low RBA unit exceeded marginal means in the high RBA and control unit by an estimated 
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[1]`
and
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[2]`
m² per hectare, respectively (p<0.01 for both, `r appNums("standFit.contrasts",display="cite")`).

The low RBA unit also significantly increased its stand density by an estimated 
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Stand density") %>%   pull(estimate) %>% round(.,0)  %>%   .[2]`
stems per hectare more than the control unit (p<0.001). The estimated rate of change in tree density did not differ significantly between the two harvested units.  The rate of change in volume in trees over 18.5cm DBH was barely significant (p=0.049) between the two harvested units, and the difference was not significant between the control and harvested units.

### Stand attributes by year
By year, mean plot volume, basal area and stand density did not differ significantly between both harvested units (i.e., p>0.05, `r appNums("standFitM.contrasts",display="cite")` and `r figNums("standPlots",display="cite")`).  Contrasts between the high-removal and control treatment units showed that volume, basal area and stand density differences were highest and significant (i.e., p<0.05) in the 1992 measurement year, and these differences diminshed over each subsequent measurement year.  For example, mean plot density in the high-removal treatment unit was approximately 600 trees per hectare less than the control in 1994 (p<0.05); by 2019, that difference had decreased to 219 trees per hectare (p>0.05).  Contrasts between the low-removal and the control treatment units showed a similar trend with basal area and stand density; the control unit had significanly higher basal area and stand density means in 1992, but the difference between the units diminished over each subsequent measurement, becoming non-significant (i.e., p>0.05) in 2019.  Plot means of volume did not differ significantly between the treatment units in any measurement year.

@@

Among treatments and throughout the stand development after treatment, the stands maintained a negative exponential (also referred to as 'reverse-J') diameter class distribution (`r figNums("diamK.plot",display="cite")`).  
Immediately after the partial harvest, the diameter class distribution of volume differed among treatments.  In the 27 years since partial harvest, volume in all three treatment units was mostly concentrated in the 22.5-27.5cm DBH class, and mostly comprised of fir (`r figNums("diamVol.plot",display="cite")`).  Over both time periods shown in the figure and all three treatments, most of the spruce volume was concentrated in DBH classes spanning 37.5 to 47.5cm.  Basal area trends were very similar to volume (not shown).

@@

A total of `r deadList %>% nrow()` trees (`r deadList %>% filter(Species=="Fir") %>% nrow()` fir, `r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce, and `r deadList %>% filter(Species=="Fd") %>% nrow()` Douglas-fir) died in `r length(unique(deadList$Plot))` plots from 1994 to 2019. The number of trees that died per plot from 1994 to 2019 increased significantly with plot basal area in 1994 (`r figNums("mort.plot",display="cite")`, p < 0.0001). Mortality was generally not attributable to a particular forest health factor, and in the majority ofcases, we could not ascertain a casual agent.  Similar linear models fitted separately for fir and spruce showed that fir mortality had a similar relationship to plot basal area, whereas spruce mortality did not (not shown).  The diameter distribution of fir mortality followed a negative exponential distribution, with mortality concentrated in smaller diameter classes (`r figNums("mortSppDiam.plot",display="cite")`).  Spruce mortality was concentrated in larger diameter classes.  By treatment unit, the control unit had the highest mortality, whereas the high-removal treatment unit had the lowest mortality (`r figNums("mortDiam.plot",display="cite")`).  The majority of mortality occured during the most recent measurement period of 2009-2019.

@@

## Uneven-aged stand management
This study describes tree- and stand-level responses to partial harvest treatments in an uneven-aged stand.  
- THere are numerous studies documenting responses of even-aged stands to partial harvest treatments such as thinning.  In even-aged stands, a common objective of thinning treatments is to re-allocate growing space and resources to the most valulable trees in the stand that have capacity to increase production.  The objectives of partial harvesting in uneven-aged stands is more often to provide a sustained production of wood over multiple generations of trees.  In this context, partial harvesting in uneven-aged stands seeks to provide adequate growing space and resources to trees across a range of sizes and densities.  By quantifying growth responses at the tree- and stand-level, this study addresses questions regarding the manipulation of structure in an uneven-aged stand to achieve ...

@@

In comparison, both partial harvest treatments stimulated a growth response in fir, compared to the control.  Although both species are similar in shade tolerance, firs are considered to be better adapted to low light conditions in the understory, allowing them to persist longer in undisturbed forest  (Antos 2000) and potentially able to respond faster to even modest changes in understory light (Dumais et al 2014, Messier et al. 1999).  Although fir growth may respond at lower levels of overstorey removal, the growth response potential of fir may be lower than spruce.  Comparative studies of spruce and fir growth rates present differing evidence regarding species differences in release potential to thinning.  In mature Engelmann spruce (*Picea engelmannii*) – subalpine fir forests in western North America, understory spruce growth rates exceeded those of fir, allowing spruce to achieve canopy position faster, including after canopy gaps were created through disturbance (Antos et al. 2000, Andrus et al. 2018).  In mature forest in central BC, spruce growth exceeded that of subalpine fir after age 80 (Eis and Craigdallie 1983).  Because spruce tends to live longer than subalpine fir, higher growth rates at later ages may allow this species to become larger than fir.  In contrast, thinning studies in spruce-fir forests in eastern North America report that balsam fir (*Abies balsamifera*) responds to release with faster height and diameter growth than co-occurring red spruce (*Picea rubens*) (Dumais and Prevost, 2007; Dumais et al. 2014, Messier et al. 1999). Other studies have found no difference in growth response between spruce and fir species after treatment (MCaughey and Smith1982).   Our study also suggests that at lower levels of basal area removal, spruce and fir growth responses are similar.

@@

### Stand-level growth responses
Over the 27-year period of this study, plots with high-removal harvest treatments accrued basal area significantly faster than the control and low-removal harvest treatment, likely due to increases in tree-level growth and stand density discussed previously.  As a result, differences in basal area between the high-removal harvest unit and the control reduced from ca. 17m2/hectare in 1992 to ca.9m2/hectare in 2019.

>>>>

---
title: "Manuscript_draft"
author: "Michael Jull and Hardy Griesbauer"
date: "10/12/2020"
output:
  word_document:
    reference_docx: wordTemplate2.docx
  html_document:
    df_print: paged
csl: cjfr.csl
bibliography: ep1162.bib
editor_options: 
  markdown: 
    wrap: 72
---

@@

-   Great: Very clear and straightforward. i.e. - log-transformed
    treatment (initial BA) was a very statistically significant
    predictor of tree response across dbh classes for both Sx and Bl.
-   Don't know what all the random effects outputs are. I am sure they
    are meaningful, but I just dont know what they mean.
-   Number of observations of 510: Should ensure it is clear that only
    trees \> 7.5 cm dbh are counted.
-   Would be good to define the dbh class limits somewhere. In Appendix
    6 table, we describe the class mid-point.
-   For Appendix 6, I assume that the null hypothesis for the ratio is
    = 1.

@@

-   I am thinking that mortality should / could be analysed as a
    frequency response - i.e. - expected vs actual distribution of
    mortality by size class and/or species. e.g. a Chi-squared or allied
    type of analysis. Null hypothesis is that mortality of stems is
    randomly distributed across different sizes and species of trees. H1
    is that it is non-random.
-   I'd suggest that we don't compare total mortality (e.g \# of stems)
    across treatment units, but rather frequency relative to initial
    number of stems (i.e. - at beginning of observation period). That
    way we normalize mortality rates relative to initial stem densities.
    Otherwise, results could be mis-leading - e.g. Control treatments
    could have more mortality in overall sph, simply because they have
    more initial stems.
-   ACTION ITEM FOR ME: I need to check my original datasets and data
    sorting iterations to see how I handled mortality observations for
    the 1992 and 1994 plot remeasurements.
-   Note for you: All of the mortality you've analysed and identified so
    far is mortality that is incremental following 92/94, so it is still
    solid analysis data, just with that qualification. I will clarify
    the 92/94 mortalilty observations for the Methods section; we did
    observe dead trees at these times, but these mortality pre-dated our
    treatments and/or first measurements. Anyway, probably not a big
    deal, but this needs examination and clarification to explain this.
    If necessary, I will quantify observations of dead trees observed on
    these years, because they are actually \> 0.

Testing for density-dependence of mortality responses... some ideas or
hypotheses.. Note that f (\_\_\_) means "a function of"...

-   MORT RATE of a size class = f (Sum of BA in that plot \> than that
    size class). Would assume asymmetric competition where only bigger
    trees exert competition on smaller trees.
-   MORT RATE of a size class = f ( Sum of BA in that plot that is
    either \> or \< that size class). Assumes symmetric competition
    where both big and small trees exert competition on their neighbors.

@@

-   See notes above about use of term "units" vs treatment (or similar).
-   Clarify effects of treatment and/or species on growth rate vs
    standing vol / BA at each measurement period.
-   A Figure 7b or equivalent could visually present changes in growth
    rate over time. (e.g. previous graphs we've developed)
-   Under Observations, \# 2 - the 2nd sentence needs work to edit and
    clarify.

@@

Last winter, as well as keeping reference copies of the initial "raw"
dataset following data entry and related stuff, I kept copies each stage
of data cleaning and prep (about 10 iterations).

We did have "dead tree" observations in the first \~ 2 year of
monitoring. The key thing to see is : if I can determine is when the
trees were dead, and/or when death was initiated, and why.\
Timing of tree death and observations of causal factors of tree death in
these first two years will help inform our interpretation and data
handling of these observations.

If trees were dead or moribund at the first "initial" observation or two
(as soon after the logging as possible), this is -- arguably - different
than mortality that occurs over time in response to the basal area
treatment and/or could be considered to be density- or
treatment-dependent in some way.

The 1992 data was in mid-March really early after logging, and before
snowmelt (in retrospect not a great idea but we had resources available
before the end of the fiscal year at the time). The 1994 data was from
when the plots were re-visited under snow-free conditions. We did a
thinning and sanitation culling of trees in the treatment units and
PSP's in 1994 to remove logging-damaged trees and most deciduous, to
bring the units and PSP's down to their final treatment BA. 1994
remeasurements were completed after that.

@@

1.  The very-early observed mortality may be a result of pre-existing
    stand conditions not the created, treatment-induced stand
    conditions. "stochastic" vs potentially treatment induced.
2.  It can be argued that trees dead at time of treatment or very soon
    thereafter were never taking up growing space in the post-harvest
    stand, and therefore would not contribute to the initial "green"
    basal area going forward.
3.  We do have 1992 data in most plots, but it was 1994 when the BA in
    each trmt unit and embedded PSP was finalized. 1994 is the most
    homogeneous start point for all PSP's.

@@

Trees that were dead at the time of the 1992 PSP establishment were not
tagged or recorded. The rationale for this was that these trees were
dead and therefore not taking up up growing space in the stands. Dead
trees in 1992 would have been either snags / dead trees pre-existing the
harvest, or trees fatally damaged or killed by logging activities (e.g.
-- trees with stems snapped and crowns broken off, no live limbs) or
trees smashed to the ground by felled trees).

@@

EP 1162 treatment units, including the PSP area contained therein, were
assessed in Spring 1994 to review stand density / basal area, and stand
condition, in order to:\
a) identify and mark-to-cut residual trees that were considered too
badly scarred or damaged to constitute good growing stock for trial
purposes, b) where needed, to lower the initial BA closer to the target
level, c) to either fell or girdle birch trees, especially in the PSP's
themselves, but usually also within the TU as a whole, and d) to fell
any dead, dangerous, and/or moribund trees within the treatment units
and PSP's. Generally most if not all of these trees felled in the
sanitation felling were understory or subdominant trees, and
occasionally a lower codominant tree. Tagged trees felled by the
sanitation felling were documented on field sheets and our Excel
datasets, so we can retrieve the species and size of these trees as
needed. However, sanitation-felled trees included both live, moribund,
and dead trees -- generally all poorer quality or vigor though.

After felling, the felled trees were bucked to a 2 to 3 m length so as
to lay these felled trees to as close to the ground as possibly,
maximize decay, and minimize physical obstructions within the TU's and
PSP's.

@@

We don't, therefore, have data on mortality between 1992 and 1994,
though my recollection is that it was minor in nature.

@@

"The difference in basal area between 1992 and 1994 is a result of basal
area reductions through the sanitation felling of some stems and some
minor natural mortality, combined with basal area increases through the
growth of remaining live stems."

A review of the various plots indicates that between 1992 and 1994, some
PSP's decreased in BA (as sanitation removals and minor mortality
exceeded growth of live stems), while others increased in BA as tree
growth exceeded BA removals.

-   Is there any change in which sizes or size classes of trees die as
    stand volume increases? - e.g. is competition asymmetrical or
    symmetrical? Currently the y axis is total number of trees, I
    believe.
-   Not sure if we have enough sample size of trees to look at species
    dynamics (Sx vs Bl).
-   I am wondering if the same relationship that appears between volume
    and tree mortality, is mirrored in the relationship between BA and
    tree mortality. A bit more on this below...

@@

To test the hypothesis that tree-level radial growth response over the
1994-2019 period varied with interactions between (i) tree diameter at
the time of final treatment (1994); (ii) treatment; and (iii) species,
we fit linear mixed effects models with log-transformed 1994-2019
diameter increment as the response variable, and treatment, species and
log-transformed 1992 diameter as interacting fixed effects. To account
for plot-level productivity differences, we included plot as a random
intercept effect. Models were fit using the 'lme4' package for R [\@].
We also fit a similar model with the 1994-2019 diameter increment
expressed as a percentage of 1994 diameter. Plots of residuals and fit
lines with orginal data were examined to evaluate model goodness of fit.
Variance explained for each model are reported using the approach
described in Nakagawa et al. [-@Nakagawa2013], as implemented in the
MuMIn package for R [@barton2009mumin]. Pairwise contrasts were
performed using the 'emmeans' package for R [\@] and Tukey HSD method of
*p* value adjustment was used to control for multiplicity. We report all
significant results at p\<0.05.

For each plot and measurement period, we computed the growth dominance statistic (GDS) to evaluate the relationship between individual tree size and its growth as a proportion of the stand's overall growth [@Forrester; @West; @Binkey]

@@

#### Stand-level volume and basal area estimates

Total and merchantable volumes were estimated for trees using species-
and region-specific equations [@nigh2016] based on tree height and DBH.
Consistent with forest management practices in British Columbia, we
estimated volumes for trees meeting a minimum diameter utilization limit
of 18.5cm DBH. For trees lacking a tree height measurement, we estimated
height by fitting a single diameter-height model developed with a linear
mixed effects model, with log-transformed DBH as a fixed effect, and
tree nested within plot as random intercepts. Trees with broken tops
were omitted from diameter-height models. We summed tree-level estimates
of volume and basal area to generate stand-level estimates for each
plot. To test the hypothesis that changes in stand attributes over 27
years varied by treatment, we fit separate linear models with stand
total volume, basal area, and density as response variables, and the
interaction between treatment and year as the independent variable.

@@

The tree-level basal area increment model indicated that interactions
between log-transformed tree size, species and treatment explained
`r round(r.squaredGLMM(fitBA)[2]*100,1)`% of the variation in
log-transformed 1994-2019 tree-level BAI, and
`r round(r.squaredGLMM(fitBApct)[2]*100,1)`% of the variation in basal
area increment as a percentage of tree size in 1994 (p\<0.0001 for both,
`r tableNums("fitBA.table",display="cite")`). Estimated marginal means
from the model showed that between species and among treatments, basal
area increment increased with tree size, meaning that larger trees
increased basal area more than smaller trees
(`r figNums("treeGrowth.plot",display="cite")`), although smaller trees
increased their size proportionally more than large trees
(`r figNums("treeGrowthPct.plot",display="cite")`). Species contrasts
showed that spruce trees increased basal area significantly more than
than fir in the high-removal and control treatment units, whereas basal
growth between the two species was close to equal in the low-removal
unit (`r appNums("sppContrast.table",display="cite")`). Within-species
contrasts showed that spruce basal area increment in the high-removal
unit exceeded spruce growth in both other units
(`r appNums("trtContrast.table",display="cite")`), whereas fir basal
area increment in the high- and low-removal units was higher than the
control unit, and there were no significant differences between the two
harvested units (`r appNums("trtContrast.table",display="cite")`).

@@

### Tree height modeling

Tree heights had a positive nonlinear relationship with diameter, for
both species and across all time periods
(`r appNums("heightDiam.plot",display="cite")`). We originally developed
separate models for each species and time period, however, a comparison
of height-diameter models showed that intercepts and slopes did not vary
significantly (i.e., p\>0.05) between species or time periods (not
shown). Therefore, we developed a single height-diameter linear
mixed-effects model using
`r X %>% filter(Species %in% c("Fir","Spruce")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()`
tree height-diameter observations in the dataset. This model explained
`r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in
log-transformed heights, with an intercept of
`r round(fixef(htDiamFit)[1],3)` and slope coefficient of
`r round(fixef(htDiamFit)[2],3)` \* log-transformed DBH (not shown).
Residual and predicted vs actual height plots were assessed to ensure
goodness of fit (`r appNums("htModelFit.plot",display="cite")`). Using
this model, we generated height estimates for trees without height
measurements, and then estimated tree total volume as per Nigh
[-@nigh2016].

@@

### Rate of change from 1994 to 2019

All treatment units, including the control, increased volume, basal area
and stand density over the 1994-2019 period
(`r figNums("standPlots",display="cite")` and
`r appNums("standPlot.table",display="cite")`). Mean annual basal area
and volume increment peaked during the 1997-2009 period in the two
harvested treatments, and peaked in the 1994-1997 period in the control
treatment unit (`r figNums("perInc.plot",display="cite")`). From 1994 to
2019, estimated marginal means of basal area increment in the low RBA
unit exceeded marginal means in the high RBA and control unit by an
estimated
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[1]`
and
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Basal area") %>%   pull(estimate) %>% round(.,1) %>% .[2]`
m² per hectare, respectively (p\<0.01 for both,
`r appNums("standFit.contrasts",display="cite")`).

The low RBA unit also significantly increased its stand density by an
estimated
`r standFit.contrasts %>% as.data.frame() %>% filter(Variable=="Stand density") %>%   pull(estimate) %>% round(.,0)  %>%   .[2]`
stems per hectare more than the control unit (p\<0.001). The estimated
rate of change in tree density did not differ significantly between the
two harvested units. The rate of change in volume in trees over 18.5cm
DBH was barely significant (p=0.049) between the two harvested units,
and the difference was not significant between the control and harvested
units.

### Stand attributes by year

By year, mean plot volume, basal area and stand density did not differ
significantly between both harvested units (i.e., p\>0.05,
`r appNums("standFitM.contrasts",display="cite")` and
`r figNums("standPlots",display="cite")`). Contrasts between the
high-removal and control treatment units showed that volume, basal area
and stand density differences were highest and significant (i.e.,
p\<0.05) in the 1994 measurement year, and these differences diminshed
over each subsequent measurement year. For example, mean plot density in
the high-removal treatment unit was approximately

`r standFitM.contrasts %>% filter(Contrast=="TU:10 - TU:Control") %>% filter(Variable=="Stand density") %>%  filter(Year==1994) %>% mutate(estimate=abs(round(estimate,0))) %>%  pull(estimate)`

trees per hectare less than the control in 1994 (p\<0.05); by 2019, that
difference had decreased to

`r standFitM.contrasts %>% filter(Contrast=="TU:10 - TU:Control") %>% filter(Variable=="Stand density") %>%  filter(Year==2019) %>% mutate(estimate=abs(round(estimate,0))) %>%  pull(estimate)`

trees per hectare (p\>0.05). Contrasts between the low-removal and the
control treatment units showed a similar trend with basal area and stand
density; the control unit had significantly higher basal area and stand
density means in 1994, but the difference between the units diminished
over each subsequent measurement, becoming non-significant (i.e.,
p\>0.05) in 2019. Plot means of volume did not differ significantly
between the treatment units in any measurement year.

@@

Among treatments and throughout the stand development after treatment,
the stands maintained a negative exponential (also referred to as
'reverse-J') diameter class distribution
(`r figNums("diamK.plot",display="cite")`).\
Immediately after the partial harvest, the diameter class distribution
of volume differed among treatments. In the 27 years since partial
harvest, volume in all three treatment units was mostly concentrated in
the 22.5-27.5cm DBH class, and mostly comprised of fir
(`r figNums("diamVol.plot",display="cite")`). Over both time periods
shown in the figure and all three treatments, most of the spruce volume
was concentrated in DBH classes spanning 37.5 to 47.5cm. Basal area
trends were very similar to volume (not shown).

@@

A total of `r deadList %>% nrow()` trees
(`r deadList %>% filter(Species=="Fir") %>% nrow()` fir,
`r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce, and
`r deadList %>% filter(Species=="Fd") %>% nrow()` Douglas-fir) died in
`r length(unique(deadList$Plot))` plots from 1994 to 2019. The number of
trees that died per plot from 1994 to 2019 increased significantly with
plot basal area in 1994 (`r figNums("mort.plot",display="cite")`, p \<
0.0001). Mortality was generally not attributable to a particular forest
health factor, and in the majority ofcases, we could not ascertain a
casual agent. Similar linear models fitted separately for fir and spruce
showed that fir mortality had a similar relationship to plot basal area,
whereas spruce mortality did not (not shown). The diameter distribution
of fir mortality followed a negative exponential distribution, with
mortality concentrated in smaller diameter classes
(`r figNums("mortSppDiam.plot",display="cite")`). Spruce mortality was
concentrated in larger diameter classes. By treatment unit, the control
unit had the highest mortality, whereas the high-removal treatment unit
had the lowest mortality (`r figNums("mortDiam.plot",display="cite")`).
The majority of mortality occured during the most recent measurement
period of 2009-2019.

@@

## Uneven-aged stand management

This study describes tree- and stand-level responses to partial harvest
treatments in an uneven-aged stand.\
- THere are numerous studies documenting responses of even-aged stands
to partial harvest treatments such as thinning. In even-aged stands, a
common objective of thinning treatments is to re-allocate growing space
and resources to the most valulable trees in the stand that have
capacity to increase production. The objectives of partial harvesting in
uneven-aged stands is more often to provide a sustained production of
wood over multiple generations of trees. In this context, partial
harvesting in uneven-aged stands seeks to provide adequate growing space
and resources to trees across a range of sizes and densities. By
quantifying growth responses at the tree- and stand-level, this study
addresses questions regarding the manipulation of structure in an
uneven-aged stand to achieve ...

@@

In comparison, both partial harvest treatments stimulated a growth
response in fir, compared to the control. Although both species are
similar in shade tolerance, firs are considered to be better adapted to
low light conditions in the understory, allowing them to persist longer
in undisturbed forest (Antos 2000) and potentially able to respond
faster to even modest changes in understory light (Dumais et al 2014,
Messier et al. 1999). Although fir growth may respond at lower levels of
overstorey removal, the growth response potential of fir may be lower
than spruce. Comparative studies of spruce and fir growth rates present
differing evidence regarding species differences in release potential to
thinning. In mature Engelmann spruce (*Picea engelmannii*) -- subalpine
fir forests in western North America, understory spruce growth rates
exceeded those of fir, allowing spruce to achieve canopy position
faster, including after canopy gaps were created through disturbance
(Antos et al. 2000, Andrus et al. 2018). In mature forest in central BC,
spruce growth exceeded that of subalpine fir after age 80 (Eis and
Craigdallie 1983). Because spruce tends to live longer than subalpine
fir, higher growth rates at later ages may allow this species to become
larger than fir. In contrast, thinning studies in spruce-fir forests in
eastern North America report that balsam fir (*Abies balsamifera*)
responds to release with faster height and diameter growth than
co-occurring red spruce (*Picea rubens*) (Dumais and Prevost, 2007;
Dumais et al. 2014, Messier et al. 1999). Other studies have found no
difference in growth response between spruce and fir species after
treatment (MCaughey and Smith1982). Our study also suggests that at
lower levels of basal area removal, spruce and fir growth responses are
similar.

@@

### Stand-level growth responses

Over the 27-year period of this study, plots with high-removal harvest
treatments accrued basal area significantly faster than the control and
low-removal harvest treatment, likely due to increases in tree-level
growth and stand density discussed previously. As a result, differences
in basal area between the high-removal harvest unit and the control
reduced from ca. 17m2/hectare in 1992 to ca.9m2/hectare in 2019.
